# Week6 - Analysis of longitudinal data

This final week we took a look at longitudinal data which contains several measurements or observations of the reponse variable of interest. For such data we first looked at graphical exploration method and simple analysis method. Then we also looked at methods that aim to account for the correlated nature of the data and where the response is assumed to be normally distributed.

## Data Wrangling

The first task was to perform some data wrangling on the raw data. Raw data was in the wide form and we needed to turn the data to long form. In the wide form the repeating measurments for one observation are as variables and when turn it in to long from by transforming those variables as new observartions.  

My data wrangling source code can be found from [here](https://github.com/ittobor/IODS-project/blob/master/data/meet_and_repeat.R).  

I saved the wrangling long data results of both of the files, and use those files here in this rmarkdown.  

## Graphical Displays and Summary Measure Approach with RATS data

First we load and take a look how the long RATS data looks like with glimpse. And we realize that ID an Group needs to be foctorized, do that and the do glimpse again. 

```{r echo=TRUE, message=FALSE}
library(dplyr)
RATS <- read.table("https://raw.githubusercontent.com/ittobor/IODS-project/master/data/RATS.txt", sep="\t", header=TRUE)
glimpse(RATS)

RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
glimpse(RATS)
```

From summary of RATS we observe that there are 3 groups and that the amount of observation in Group1 is larger that in the other 2 groups.

```{r echo=TRUE, message=FALSE}
summary(RATS)
```

Then we plot the RATS dataset that has data from nutrition study for rats, with 3 groups getting different diet and follow-up on their weight in course of time. We plot values for all the rats differentiating by between the groups and track the weight of the individual rats.

We can see that during the study in Group1 all the 8 rats weight almost the same and all are under 300 (g) which at the same time is clearly less than in the other groups. There is one rat in Group one that little bit stands out from the rest by weighting under 250 (g) for the durance of the study.  

In Group2 there are 4 rats with 3 of them with similar weight (400-500 g) during the study. One rat stands out by outweighting the others  (550-650 g).

In Group3 there are 4 rats with 3 of them with similar weight (500-600 g) during the study. One rat stans out by weighting less than the other (450-55 g).

```{r echo=TRUE, message=FALSE}
library(ggplot2)
ggplot(RATS, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATS$Weight), max(RATS$Weight)))

```

To see the tracking more clearly we can standardize the data and then plot again. We can see some change in the slopes of the lines, as they tend to get less steep some even goes negative. We can now better compare how the weight changes of the individuals compared to the mean weight. The same individuals still stand out but that is expected.  

With nutrition in Group1 the weight change of the individuals seem steady when comparing to others. There is on individual whos weight first drops to some level but then stays around that quite stable.

With nutrition in Group2 there is one indidual whos weight is not effected that much when compared to others, the slope is downwards meaning other individuals are gaining weight relatively much more. Other individuals in this group are reletively gaining more weight the individuals in all groups.

With nutrition in Group3 3 out of 4 individuals are gaining weight relatively less than other in all groups, the slope seems downwards for them. There is 1 individual whos weight gain is slighly upwards.

```{r echo=TRUE, message=FALSE}
RATSL <- RATS %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(name = "standardized Weight")
```

Next we combine the individuals in groups and at the average profiles of each of the nutrition groups with standard error mean as indicator of variance. There is no overlapping of the curves nor the variances, then again all the starting weight of the individuals between the groups were on different levels.  

Group1 seems to have the smallest variance, ~5g~,and the weight gainded on average is smallest, around 25g during the study period.  

Group2 has the biggest variance, ~ 50g, and the weight gained is the biggest, around 75g durding the study period.  

Group3 has the second bigget variance. ~ 20g, and the weight gained is the second best, around 40g during the study period.

```{r echo=TRUE, message=FALSE}
n <- RATSL$Time %>% unique() %>% length()
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

#glimpse(RATSS)

ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = "top") +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

Alternative way to compare the avarages of the three groups is side-by-side boxplot below. I removed the the first day observastion, eventhough it does not make any difference for the results. But there is linear regression use case where it might effect the results. From the boxplot we can see the existance of outliers in each of the groups.

```{r echo=TRUE, message=FALSE}
RATSLS <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

#glimpse(RATSLS)

ggplot(RATSLS, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), whole study period")
```

Then let's filter out the outliers in each group and plot again. By comparing the two sets of boxplots (above and below). So even with the boxplots we can see that the groups 2 and 3 look similar, but then again they all operate on different levels of weight.

```{r echo=TRUE, message=FALSE}
# one method to remove the outliers
RATSLSG3 <- RATSLS %>% filter(Group == 3)
RATSLSG3 <- RATSLSG3 %>% filter(mean > 500)
RATSLSGR <- RATSLS %>% filter(Group != 3)
RATSLSGR <- RATSLSGR %>% filter(mean > 250) %>% filter(mean < 550)
RATSLSR <- rbind(RATSLSGR, RATSLSG3)

ggplot(RATSLSR, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), whole study period")
```

Next we perform t-test to see any difference between the groups. T-test only applies with 2 groups so with 3 groups we need to compare 3 different pairs.  

There is big difference in the the means with Group1 towards both groups 2 and 3. The p-value between Group1 and both groups 2 and 3 is in the significant level. Also 0 in not in the confidence interval in either of the group comparisons between Group1 to groups 2 and 3. So there is evidence that Group1 differs from both groups 2 ans 3.  

On the contrast the comparison bethween Group2 and Group3 provides evidence that they are not diffent. This because p-value in higher that the significance level, 0 is included in the confidence interval and mean values of the groups are not that far form each other after all.

```{r echo=TRUE, message=FALSE}
RATSLSRg12 <- RATSLS %>% filter(Group != 3)
t.test(mean ~ Group, data = RATSLSRg12, var.equal = TRUE)
RATSLSRg13 <- RATSLS %>% filter(Group != 2)
t.test(mean ~ Group, data = RATSLSRg13, var.equal = TRUE)
RATSLSRg23 <- RATSLS %>% filter(Group != 1)
t.test(mean ~ Group, data = RATSLSRg23, var.equal = TRUE)
```

Lastly we will use Weight value corresponding to time (WD1) at the start of the nutrition study as the baseline covariate. And do this for all the Groups in the data. We can see from the result that the baseline Weight is strongly related to values taken after study started and also there is mild evidence of Group difference when conditioning on the baseline value.

```{r echo=TRUE, message=FALSE}
RATSLSU <- RATSLS
RATSLSU$baseline <- as.numeric(RATS$Weight[1:16])

fit <- lm(mean ~ baseline + Group, data = RATSLSU)
anova(fit)

```

## Linear Mixed Effects Models for Normal Response Variables with BPRS data

This time we load and take a look how the long BPRS data looks like with glimpse. The `subject` values are overlapping in the `treatment` groups and we want to distinguish those. Therefore we add 20 to is `subject` value in the second `treatment`group to make the difference. ALso here we realize that ID an Group needs to be foctorized, do that and the do glimpse again.

```{r echo=TRUE, message=FALSE}
BPRS <- read.table("https://raw.githubusercontent.com/ittobor/IODS-project/master/data/BPRS.txt", sep="\t", header=TRUE)
glimpse(BPRS)
BPRSt1 <- filter(BPRS, treatment == 1)
BPRSt2 <- filter(BPRS, treatment == 2)
BPRSt2$subject <- BPRSt2$subject + 20
BPRS <- rbind(BPRSt1,BPRSt2)
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
summary(BPRS)
```

With fixing the issues with duplicate `subject` values then we can plot all individual in both `treatment` groups to the diagram. From the diagram it is not practical to try to infer anything other than the similarity between the groups.

```{r echo=TRUE, message=FALSE}
ggplot(BPRS, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = treatment)) + 
  scale_x_continuous(name = "week (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
```

Next we fit multiple linear regression model with `bprs` as the response and `week` and `treatment` as explanatory variables. We can see the significance of `week` and insignificance of `treatment` group 2 and absence of `treatment` group 1. The result does not change remarkably if the initial Week 0 is not included. We note here that this model assumes independence of the repeated measurements.

```{r echo=TRUE, message=FALSE}
BPRS_reg <- lm(bprs ~ week + treatment, data=BPRS)
summary(BPRS_reg)
```

Since the repeated measurements of bprs are unlikely independent, we will fit more appropriate model: random intercept model. Here we use `week` and `treatment` as fixed-effect terms and `1 | subject` as random-effect term. 

```{r echo=TRUE, message=FALSE}
library(lme4)
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data=BPRS, REML=FALSE)
summary(BPRS_ref)
```

Next we fit random intercept and random slope model with `week | subject` as the random intercept and slope, and we compare it to the previous random intercept model with analysis of variances. We can see from the chi-squared value and p-value that with lower values the random intercept and random slope model performs better and has true significance.

```{r echo=TRUE, message=FALSE}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRS, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

We will fit yet another random intercept and random slope model by adding `week | treatment` to the model as the random intercept and slope, and we compare the new model to the previous model with analysis of variances. In my interpretation the latest addition does not increase the significance form the previous model.

```{r echo=TRUE, message=FALSE}
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject) + (week | treatment), data = BPRS, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)
```

Here we will plot the diagram of BPRS with fitted `bprs` values from with both andom intercept and random slope models. The diagram of BPRS with observed `bprs` values can be found from above, but it is plotted as the last plot again so it is easier to compare. From fitted ones we can see they don't seem to differ at all, as was anticipated.

```{r echo=TRUE, message=FALSE}
Fitted <- fitted(BPRS_ref1)
BPRS$Fitted <- Fitted
ggplot(BPRS, aes(x = week, y = Fitted, group = subject)) +
  geom_line(aes(linetype = treatment)) + 
  scale_x_continuous(name = "week (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top") +
  ggtitle("BPRS_ref1 fitted")

Fitted <- fitted(BPRS_ref2)
BPRS$Fitted <- Fitted
ggplot(BPRS, aes(x = week, y = Fitted, group = subject)) +
  geom_line(aes(linetype = treatment)) + 
  scale_x_continuous(name = "week (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top") +
  ggtitle("BPRS_ref2 fitted")

ggplot(BPRS, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = treatment)) + 
  scale_x_continuous(name = "week (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")  +
  ggtitle("BPRS observations")
```