# Week3 - Logistic Regression

I prepared myself for this weeks topic by completing DataCamp exercise on "Logistic Regression".

## Data Wrangling
As I did last week, for the data wrangling part of this weeks exercises I mostly used the methods learned from the DataCamp exercise. I learned about joining datasets and doing data mutations.

My source code used in the wrangling is [here](https://github.com/ittobor/IODS-project/blob/master/data/create_alc.R).  

Basically, here are the main points what is done in the code:

* Initial dataset is acquired and extracted manually from given web source.
* Two data sets are read in and combined as one, since they contain answers from same students, by joining them based on variables: `school`,`sex`,`age`,`address`,`famsize`,`Pstatus`,`Medu`,`Fedu`,`Mjob`,`Fjob`,`reason`,`nursery`,`internet`
* The columns not used in join are combined by taking rounded avarage of numerical values and by selecting the first value of not numeric values
* `alc_use` variable was generated to indicate alcohol use (avarage of daily and weekly consumption)
* `high_use` variables was generated to indicate high use of alcohol (true is `alc_use` is > 2)

The dataset (`alc.csv`) that the code created can be found from c

## Data Analysis

### Step 1 - New markdown file
Step one was to create chapter3.Rmd to inject this content to the diary as a child of subscipts in index.Rmd.

### Step 2 - Combined dataset and attributes 

The combined dataset contains the data from two Portuguese schools and were originally gathered from school reports and questionnaires. The data was collected initally in two datasets, one regarding student performance in mathemathics and one regarding student performance in Portuguese language. In addition to section Data Wrangling, in which the join criteria and additionally created variables are discussed, you can find more information about the variables in [here](https://archive.ics.uci.edu/ml/datasets/Student+Performance).

```{r echo=FALSE, message=FALSE}
library(ggplot2)
library(GGally)
library(dplyr)
library(tidyr)
alc <- read.table("https://raw.githubusercontent.com/ittobor/IODS-project/master/data/alc.csv", sep=";", header=TRUE)
```
Dimensions and the column (variable) names of the combined dataset:
```{r echo=TRUE}
dim(alc)
colnames(alc)
```
Column (variable) types of the combined dataset:
```{r echo=TRUE}
glimpse(alc)
```

### Step 3 - Variables to study in relations with alcohol consumption

My hypothesis to study is how time consumption is related with alcohol consumption. The idea being that if you spend your time with activities you would not have time to consume alcohol and vice versa. My choice of variables are in sense time consuming variables: `freetime`, `goout`, `studytime` and `traveltime`.  

* `traveltime` - Home to school travel time. My intuition here is this would be a factor to decrease alcohol consumption.
* `studytime` - Weekly study time. My intuition here is this would be a factor to decrease alcohol consumption.
* `freetime` - Free time after school. My intuition here is this would be a factor to increase alcohol consumption.
* `goout` - Going out with friends. My intuition here is this would be a factor to increase alcohol consumption.

### Step 4 - Numerical and graphical exploration on chosen variables
```{r echo=FALSE}
alc_min <- data.frame(matrix(ncol=0,nrow=382))
alc_min$high_use <- alc$high_use
alc_min$traveltime <- alc$traveltime
alc_min$studytime <- alc$studytime
alc_min$goout <- alc$goout
alc_min$freetime <- alc$freetime
ggpairs(alc_min, 
        mapping = aes(col = high_use, alpha = 0.3), 
        lower = list(combo = wrap("facethist", bins = 20)))
```
